strucplots---Visualizing higher-dimensional Contingency Tables in \pkg{vcd}

### 2005-08-04 trunk/vcd/inst/doc/strucplot.Rnw
Vignette Nr. 2 added. Temp files moved to pics subdir, which has been added to .Rbuildignore
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@421 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
sieveplot added to the strucplot framework; old function removed. line_col argument added to shading functions for default line color. dimnames added to independence_table.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@431 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine=R,eps=TRUE}
\setkeys{Gin}{width=0.8\textwidth}

@@

The \code{strucplot} framework in \pkg{vcd} can be used to visualize
higher-dimensional contingency tables and integrates techniques such as
mosaic displays and association plots. The main idea is to visualize
the tables' cells arranged in rectangular form. To allow for more than
two dimensions, the plot area is constructed using recursive
conditional splits, given the tiles representing the levels of the
previous dimension.
This principle defines a graph `family' with still many parameters
left such as:

@@

The strucplot framework is highly modularized; Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
The core function is \code{strucplot()} which processes the parameters,
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates panel, labeling, shading, and spacing functions to produce 
the plot. The actual workhorse is the chosen panel function: currently, 
low-level panel functions for association plots (\code{struc\_assoc}) and 
mosaic plots (\code{struc\_mosaic}) are available. In addition, there are
high-level wrapper functions such as \code{mosaic()} and \code{assoc()} 
to \code{strucplot()} that conveniently set the 
parameters---it is these high-level functions that the user normally would like to use, 
although \code{strucplot()} is fully functional on its own. \code{mosaic()} and 
\code{assoc} are generic functions: for both, there exist a formula method and a default 
method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
\code{cotabplot}, are on even higher level and make use of 
code{mosaic()} and \code{assoc()} to produce plots.

@@

\begin{center}
<<Expected,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):
\begin{Sinput}
> assoc(Titanic, compress = FALSE)
\end{Sinput}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<afterplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<changeplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
assoc(Titanic)
x = Titanic
x[] = NA
x["1st", "Female", "Adult", "Yes"] = "Red"
assoc(Titanic, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}

>>>>

\SweaveOpts{engine=R,eps=FALSE,prefix.string=pics/strucplot}
\setkeys{Gin}{width=0.8\textwidth}

@@

The \code{strucplot} framework in \pkg{vcd} can be used to visualize
higher-dimensional contingency tables and integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. To allow for more than
two dimensions, the plot area is constructed using recursive
conditional splits, given the tiles representing the levels of the
previous dimension.
This principle defines a graph `family' with still many parameters
left such as:

@@

The strucplot framework is highly modularized; Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
The core function is \code{strucplot()} which processes the parameters,
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates panel, labeling, shading, and spacing functions to produce 
the plot. The actual workhorse is the chosen panel function: currently, 
low-level panel functions for association plots (\code{struc\_assoc}), 
mosaic plots (\code{struc\_mosaic}), and sieve plots (\code{struc\_sieve})
are available. In addition, there are
high-level wrapper functions such as \code{mosaic()}, \code{assoc()},
and \code{sieve()} to \code{strucplot()} that conveniently set the 
parameters---it is these high-level functions that the user normally would like to use, 
although \code{strucplot()} is fully functional on its own. \code{mosaic()}, 
\code{assoc}, and \code{sieve()} are generic functions: 
for both, there exist a formula method and a default 
method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
\code{cotabplot}, are on even higher level and make use of 
code{mosaic()}, \code{assoc()}, and \code{sieve()} to produce plots.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Expected,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to directly compare observed and expected values,
we can use a sieve plot (see Figure \ref{fig:sieve}):
\begin{Sinput}
> sieve(Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(Titanic)
@
\caption{Sieve plot for the Titanic data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent The next step is to directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):
\begin{Sinput}
> assoc(Titanic, compress = FALSE)
\end{Sinput}

@@

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<afterplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)

@@

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<changeplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
assoc(Titanic)
x = Titanic
x[] = NA
x["1st", "Female", "Adult", "Yes"] = "Red"
assoc(Titanic, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}
